\chapter{Conclusions and Future Work}
\epigraph{``Another quote.''}{\vspace{10pt}Another author}
\label{chapter:conclusions-fw}

\newpage

This chapter presents the final conclusions obtained through the research effort and the future work that can be derived.

\section{Conclusions}

\begin{itemize}

    \item Three sequence-level \acrshort{uq} methods for Spanish \acrshort{asr}, \acrshort{ts}, \acrshort{mcd}, and \acrshort{lmcd}, were tuned under a common \acrshort{cmaes} protocol and compared. \acrshort{lmcd} produces uncertainty scores that track the \acrshort{wer} substantially better than the two baselines and is the only method separated from them by a statistically significant margin on both the calibration and the test splits.

    \item Tuning on equal footing was instrumental in reaching this conclusion. The dropout-based methods only become informative at dropout rates far below the usual $0.1$--$0.2$ defaults, which underscores how sensitive the quality of an uncertainty estimate is to a single hyperparameter and how misleading an untuned comparison would have been.

    \item \acrshort{lmcd} and \acrshort{mcd} remain stable when moving from calibration to test data, whereas \acrshort{ts} degrades the most; the temperature is tied to the calibration distribution and is therefore the most fragile under distribution shift.

    \item At a correlation of $R \approx \lmcdTestR$ with the \acrshort{wer}, the \acrshort{lmcd} score is practically useful for flagging likely-erroneous transcriptions for human review or for driving selective transcription, at the cost of the extra inference required by the stochastic passes.

    \item The aspirational target of a Pearson coefficient of $0.9$ was not reached, but the main objective of improving on the baseline of \cite{rodriguez2024uncertainty} by at least \qty{20}{\percent} was met, with \acrshort{lmcd} improving the \acrshort{mcd} baseline by roughly \qty{45}{\percent} on the test split.

\end{itemize}


\section{Future Work}

\begin{itemize}
    \item Validate these findings on larger corpora, additional languages, and larger Whisper variants, to confirm that the ranking and the very low optimal dropout rates generalize beyond the current setting.

    \item Explore cheaper tuning strategies for the dropout-based methods and benchmark \acrshort{cmaes} against other tuners such as random \cite{bergstra2012random} and Bayesian \cite{snoek2012practical} search.

    \item Align the reported transcription across methods so that the \acrshort{wer} is computed on the same prediction for every method, strengthening direct comparison.

    \item Extend the evaluation to other agreement-based tasks, such as machine translation and open-ended generation, where the same distance-based scoring may apply.

    \item \todo{Future thread: propose a dedicated novel clear-box method to close the remaining gap to $0.9$, propose one or more reliability metrics beyond Pearson correlation, and evaluate \acrfull{fde} alongside the current methods.}
\end{itemize}
